### 2015-04-20 Paper1rev2/paper1rev2.tex
updated for rev2

\subsection*{Intent Extraction: Deducing the Desired Trajectory}
We first demonstrate the \textit{intent extraction} approach in human motor control, but we later show that the process is applicable to any controlled process with an invertible highest order plant term. The process begins by presuming a model of the controller. Here, we choose the well-known motion control structure of Shadmehr and Mussa-Ivaldi\cite{shadmehr1994adaptive} where the feedforward aspect of the controller perfectly predicts the plant and linearizes the system through cancellation. Additionally, linear feedback rejects position and velocity error. This model was chosen to help illuminate the approach, but later we present how to generalize this approach to more complex models without loss of generality. The equation governing the passive planar dynamics (plant) of musculoskeletal structure is of the form,
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripetal}}}_{\text{Plant}}+E=0
\end{equation}
where $M$ is the mass matrix, $q$ is the joint angles, $\dot{q}$ is joint angular velocity, $\ddot{q}$ is joint angular acceleration, $G$ contains both Coriolis and centripetal effects, and $E$ is any externally-applied torque. The motion behavior changes with the addition of feedforward and/or feedback controllers,
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripetal}}}_{\text{Plant}}+E=\underbrace{\overbrace{\hat{M}(q_d)\ddot{q}_d}^{\text{Inertia}}+\overbrace{\hat{G}(q_d,\dot{q}_d)}^{\text{Coriolis, Centripetal}}+\hat{E}}_{\text{Feedforward Controller}}+\underbrace{K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})}_{\text{Impedance, Feedback Controller}}
\end{equation}
where terms with hats over them ($\hat{M}$, $\hat{G}$, or $\hat{E}$) indicate that they represent the nervous system's best estimate of the forces and dynamics it will encounter, which is also known as an internal model \cite{shadmehr1994adaptive}. This portion of the system serves as an inverse-dynamics feedforward controller that cancels out the dynamics of the arm in the torque balance. If the nervous system has sufficient experience and is expecting $E$, it is included as part of the internal model, $\hat{E}$; otherwise $\hat{E}$ is set to zero. $K_p$ and $K_d$ are the lumped impedance and feedback terms that employ a moving state equilibrium to accomplish the desired trajectory, $q_d$. This $q_d$ has a dual meaning in that it signifies both the unknown desired trajectory that we seek to discover and also the moving equilibrium trajectory of the arm.

For typical dynamic simulations, in order to determine the trajectory of the system (i.e., the forward dynamics problem), this second-order differential equation is solved by numerical integration to determine the solution to the initial value problem in time. This entails algebraic manipulation to solve for $\ddot{q}$, followed by integration to determine the state trajectory. Intent determination takes the novel approach of instead solving for $\ddot{q}_d$,
\begin{equation}
\ddot{q}_d=\hat{M}(q_d)^{-1}\left\{M(q)\ddot{q}+G(q,\dot{q})+E-[\hat{G}(q_d,\dot{q}_d)+\hat{E}+K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})]\right\}
\end{equation}  
such that $\ddot{q}_d$ can then be integrated numerically using a differential equation solver to determine the intended state trajectory $q_d(t)$. This relies on many assumptions: The model of plant and controller must be accurate and precise. The initial conditions must be available and accurate. The mass matrix estimate, $\hat{M}$, must be invertible. Externally-applied force must be precisely and accurately measured. If all of these conditions are met, then the system yields an accurate estimate of the intent.

@@

\subsection*{Dynamic simulation of arm and intended trajectories}
While the derivation of intent determination is general, testing its application to human reaching requires choosing plant and actuator models, which themselves require physical parameters. Appropriate plant modeling is well-understood: measured arm segment lengths and self-reported body mass are converted into the plant's inertial, centripetal, and Coriolis terms. Anatomical landmarks and values from Dempster \cite{dempster1955space} and Winter \cite{winter2009biomechanics} relate body mass to limb mass, limb length to limb center of mass, and limb mass and length to moment of inertia. For the actuator, we choose the model and parameters of Burdet et al. \cite{burdet2006stability}. We fit a subject-specific constant scale factor for impedance, $c$, relative to this model to account for any task-dependence of the subjects' impedance\cite{gomi1998task}.
\begin{equation}
c\big(K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})\big)=\big(\hat{M}(q_d)\ddot{q}_d+\hat{G}(q_d,\dot{q}_d)\big)-\big(M(q)\ddot{q}+G(q,\dot{q})\big)+E
\end{equation}
This uses linear regression, as in Franklin et al.\cite{franklin2003adaptation}, except that task-perpendicular torque and displacement from the first 150 ms of movement from noise-disturbed trials is used in place of a stationary task.

Desired trajectory in time was idealized as a typical minimum jerk, 5th order polynomial of duration 700 ms starting and ending with zero velocity and acceleration \cite{flash1985coordination}.  In this  model, musculoskeletal stiffness was linearly related to torque. Like Burdet et al. \cite{burdet2006stability}, we approximated muscle torque by looking backward in time $500 \mathrm{\mu s}$. Note that this backward look led to negligible (sub-micrometer mean unsigned error) discrepancies in subsequent intent estimates.

@@

It is important to mention that it is not necessary for the brain to represent intent explicitly to generate action. It is possible that the nervous system can simply learn the relationship between muscle activations and the accomplishment of goals. For example, one might simply evaluate a motion outcome and adjust descending signals to the muscles. At the same time, one may not deny the fact that there are tasks that require a explicit representation of a trajectory, such as performing a dance, drawing a picture, or conducting an orchestra. For these tasks, intent is likely represented at  some level in the nervous system. Nevertheless, the mechanics of the arm dictate the existence of an equilibrium whether intent is explicitly represented or not. Here our goal was to answer a more modest question, ``where would the arm have gone?''

@@

\section*{Figure Legends}
% This section is for figure legends only, do not include
% graphics in your manuscript file.
%
%\begin{figure}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure caption.  
%}
%\label{Figure_label}
%\end{figure}
\begin{figure}[!b]
\includegraphics{fig1.eps}
\caption{
{\bf Simulated data illustrating tautology of extraction across pulse and filtered Gaussian noise disturbance types.} Intent is modeled as a minimum jerk, 5th order polynomial. Forces experienced are combined with intent via Burdet et al.'s \cite{burdet2006stability} model to produce the simulated arm trajectory. Extraction to recover intention from arm and force trajectory follows. Parameter errors are introduced into the extraction and varied according to Table 1 to estimate sensitivity. Examining the distribution of deviation in trajectories from this analysis reveals that sensitivity is not uniformly distributed throughout a reach.
}
\label{fig:synthetic}
\end{figure}

\begin{figure}
\includegraphics{fig2.eps}
\caption{
{\bf Model sensitivity testing (from the variance-based method of Saltelli et al.\cite{saltelli2010variance}) reveals that the model is mostly sensitive to stiffness.} Parameter values are varied in a Sobol-distributed fashion according to Table 1. First order sensitivity, in purple, is the error that would be removed if the parameter was fixed at its nominal value. Total sensitivity, in teal, shows error that would remain if all other parameters were fixed at their nominal values.
}
\label{fig:sensitivity}
\end{figure}

\begin{figure}
\includegraphics{fig3.eps}
\caption{
{\bf Subjects' disturbed hand trajectories and extraction of intended trajectories from them reveal that both variance and invariance of the motor plan can occur even in response to very large disturbances.} The hand path, in black, deviates from the blue baseline, as a force pulse, gray arrows, is applied. Three of the eight subjects' intent trajectories (orange) did not significantly deviate from undisturbed movement (panel A) as observed by the intent region not departing the undisturbed movement (blue) region. Shaded regions represent mean $\pm$ 95\% confidence interval on the mean. The first time at which the hand (black) region departed from undisturbed movement is marked with a green dot for each subject. Five of the eight subjects' intent trajectories did significantly deviate from undisturbed movement (panel B). The time of intent departure from the undisturbed trajectories is marked with a red dot for each subject.
}
\label{fig:anecdotes}
\end{figure}
\pagebreak

>>>>

\subsection*{Intent Extraction: Deducing the Desired Trajectory}
We first demonstrate the \textit{intent extraction} approach in human motor control, but we later show that the process is applicable to any controlled process with an invertible highest order plant term. The process begins by presuming a model of the controller. Here, we choose the well-known motion control structure of Shadmehr and Mussa-Ivaldi\cite{shadmehr1994adaptive} where the feedforward aspect of the controller perfectly predicts the plant and linearizes the system through cancellation. Additionally, linear feedback rejects position and velocity error. This model was chosen to help illuminate the approach, but later we present how to generalize this approach to more complex models without loss of generality. The equation governing the passive planar dynamics (plant) of musculoskeletal structure is of the form,
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripetal}}}_{\text{Plant}}+E=0
\end{equation}
where $M$ is the mass matrix, $q$ is the joint angles, $\dot{q}$ is joint angular velocity, $\ddot{q}$ is joint angular acceleration, $G$ contains both Coriolis and centripetal effects, and $E$ is any externally-applied torque. The motion behavior changes with the addition of feedforward and/or feedback controllers,
\begin{equation} \label{eq:controlLaw}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripetal}}}_{\text{Plant}}+E=\underbrace{\overbrace{\hat{M}(q_d)\ddot{q}_d}^{\text{Inertia}}+\overbrace{\hat{G}(q_d,\dot{q}_d)}^{\text{Coriolis, Centripetal}}+\hat{E}}_{\text{Feedforward Controller}}+\underbrace{K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})}_{\text{Impedance, Feedback Controller}}
\end{equation}
where terms with hats over them ($\hat{M}$, $\hat{G}$, or $\hat{E}$) indicate that they represent the nervous system's best estimate of the forces and dynamics it will encounter, which is also known as an internal model \cite{shadmehr1994adaptive}. This portion of the system serves as an inverse-dynamics feedforward controller that cancels out the dynamics of the arm in the torque balance. If the nervous system has sufficient experience and is expecting $E$, it is included as part of the internal model, $\hat{E}$; otherwise $\hat{E}$ is set to zero. $K_p$ and $K_d$ are the lumped impedance and feedback terms that employ a moving state equilibrium to accomplish the desired trajectory, $q_d$. This $q_d$ has a dual meaning in that it signifies both the unknown desired trajectory that we seek to discover and also the moving equilibrium trajectory of the arm.

For typical dynamic simulations, in order to determine the trajectory of the system (i.e., the forward dynamics problem), this second-order differential equation is solved by numerical integration to determine the solution to the initial value problem in time. This entails algebraic manipulation to solve for $\ddot{q}$, followed by integration to determine the state trajectory. Intent determination takes the novel approach of instead solving for $\ddot{q}_d$,
\begin{equation} \label{eq:solveDDQ}
\ddot{q}_d=\hat{M}(q_d)^{-1}\left\{M(q)\ddot{q}+G(q,\dot{q})+E-[\hat{G}(q_d,\dot{q}_d)+\hat{E}+K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})]\right\}
\end{equation}  
such that $\ddot{q}_d$ can then be integrated numerically using a differential equation solver to determine the intended state trajectory $q_d(t)$. This relies on many assumptions: The model of plant and controller must be accurate and precise. The initial conditions must be available and accurate. The mass matrix estimate, $\hat{M}$, must be invertible. Externally-applied force must be precisely and accurately measured. If all of these conditions are met, then the system yields an accurate estimate of the intent.

@@

\subsection*{Dynamic simulation of arm and intended trajectories}
While the derivation of intent determination is general, testing its application to human reaching requires choosing plant and actuator models, which themselves require physical parameters. Appropriate plant modeling is well-understood: measured arm segment lengths and self-reported body mass are converted into the plant's inertial, centripetal, and Coriolis terms. Anatomical landmarks and values from Dempster \cite{dempster1955space} and Winter \cite{winter2009biomechanics} relate body mass to limb mass, limb length to limb center of mass, and limb mass and length to moment of inertia. For the actuator, we choose the model and parameters of Burdet et al. \cite{burdet2006stability}. We fit a subject-specific constant scale factor for impedance relative to this model to account for any task-dependence of the subjects' impedance\cite{gomi1998task} as described in the next section. Desired trajectory in time was idealized as a typical minimum jerk, 5th order polynomial of duration 700 ms starting and ending with zero velocity and acceleration \cite{flash1985coordination}.  In this  model, musculoskeletal stiffness was linearly related to torque. Like Burdet et al. \cite{burdet2006stability}, we approximated muscle torque by looking backward in time $500 \mathrm{\mu s}$. Note that this backward look led to negligible (sub-micrometer mean unsigned error) discrepancies in subsequent intent estimates.

\subsection*{Torque Calculations}
The impedance properties of the human arm are known to be task-dependent\cite{gomi1998task}, but we need to model them in the context of our task without assuming our conclusions in the premises. To accomplish this, we fit a linear scale factor to the torque output an established model using an intermittently, unpredictably presented secondary task. This unpredictability enables us to leverage the ergodic assumption: our tasks should at least initally share the same motor plan and impedance properties. Both portions of this assumption are important. First, we fit some expectation of the motor plan, $x_d$, by using a Gaussian weighted average ($\sigma=22$ milliseconds) of each subject's hand trajectory as it evolves in time during undisturbed movement. Separate expectations were fit for each pair of movement start and end points. We then use inverse kinematics to convert these expected hand trajectories to expected joint angle trajectories, $q_d$. Taking time derivates of $q_d$ then allows us to form an expectation of the feedforward torque, $\tau_{ff}$, that will be present early in movement as in \eqref{eq:controlLaw},
\begin{equation}
\tau_{ff}=\hat{M}(q_d)\ddot{q}_d+\hat{G}(q_d,\dot{q}_d)+\hat{E}
\end{equation}
which then allows us to construct the full torque balance for the joints,
\begin{equation} \label{eq:torqueBalance}
\tau_P+E=\tau_{ff}+c\tau_{fb}
\end{equation}
where $\tau_P$ is calculated from the arm's dynamics, $\tau_{fb}$ is calculated from the impedance model of Burdet et al.\cite{burdet2006stability}, and $c$ is the scaling parameter to be fit. We calculate $\tau_P$ using measured limb segment lengths, $L_1$ and $L_2$ and self-reported body mass, $m_g$. We converted from these gross measurements to specific parameters using the nominal ratios provided in Table~\ref{tab:parameters},
\begin{align}
m_1 &=0.028 m_g & L_{m1}&=0.426 L_1 & J_1&=m_1(0.322 L_1)^2 \\  
m_2 &=0.022 m_g & L_{m2}&=0.682 L_2 & J_2&=m_2(0.468 L_2)^2
\end{align}
where $m_1$ and $m_2$ are segment masses, $L_{m1}$ and $L_{m2}$ are limb centers of mass, and $J_1$ and $J_2$ are mass moments of inertia. We then calculate $M$, $G$, and $\tau_P$ using these subject-specific parameters and measured values for the joint trajectory, $q$, and its time derivatives from reaches that received the white noise force disturbance,
\begin{align}
M&=\begin{bmatrix}J_1+J_2+m_1 L_{m1}^2+m_2(L_1^2+L_{m2}^2+2 L_1 L_{m2} \cos{q_2}) & J_2+m_2(L_{m2}^2+L_1 L_{m2} \cos{q_2}) \\
J_2+m_2(L_{m2}^2+L_1 L_{m2} \cos{q_2}) & J_2+m_2L_{m2}^2
\end{bmatrix} \\
G&=\begin{bmatrix}
m_2 L_1 L_{m2} \dot{q}_2(2\dot{q}_1+\dot{q}_2)\sin{q_2} \\
m_2 L_1 L_{m2} \dot{q}_1^2\sin{q_2}
\end{bmatrix} \\
\tau_P&=M(q)\ddot{q}+G(q,\dot{q})
\end{align}
for the first 150 milliseconds following the onset of movement. $\tau_{ff}$ was calculated in the same fashion $\tau_{ff}$ except that $\hat{M}$ and $\hat{G}$ were calculated by replacing all instances of $q$ with $q_d$. We then calculated $\tau_{fb}$ as in Burdet et al.\cite{burdet2006stability} while additionally leveraging the dual identity of the muscle torque, $\tau_m$,
\begin{align}
|\tau_m|&=|\tau_{ff}+\tau_{fb}|=|\tau_P+E| \\
K&=\begin{bmatrix}
10.8+3.18|\tau_{m1}| & 2.83+2.15|\tau_{m2}| \\
2.51+2.34|\tau_{m2}| & 8.67+6.18|\tau_{m2}|
\end{bmatrix}\\
\tau_{fb}(t)&=K\big(q_d(t)-q(t)+\frac{\dot{q}_d(t)-q(t)}{12}+\frac{q_d(t-\phi)-q(t-\phi)+2(\dot{q}_d(t-\phi)-q(t-\phi))}{50}\big)
\end{align}
where $\phi$ is a delay of 60 milliseconds. Finally, constrained optimization is performed to find the value of $c$ that minimizes the expression,
\begin{equation}
\tau_{ff}+c\tau_{fb}-\tau_P-E
\end{equation}
for each subject as the sum of all torques acting on the joints should ideally be zero. The optimization is constrained such that $c>0.15$ both to avoid unphysiological values and numerical instabilities during simulation. The value of $c$ is then used when extracting intended trajectories from pulse-disturbed movements as in \eqref{eq:solveDDQ},
\begin{equation}
\ddot{q}_d=\hat{M}(q_d)^{-1}\left\{M(q)\ddot{q}+G(q,\dot{q})+E-[\hat{G}(q_d,\dot{q}_d)+c\tau_{fb}]\right\}
\end{equation}
and integrated numerically as described. This provides a check on the fit value $c$ as an inappropriate value would cause $q_d$ to deviate from reasonable values.

@@

It is important to mention that it is not necessary for the brain to represent intent explicitly to generate action. It is possible that the nervous system can simply learn the relationship between muscle activations and the accomplishment of goals. For example, one might simply evaluate a motion outcome and adjust descending signals to the muscles. At the same time, one may not deny the fact that there are tasks that require a explicit representation of a trajectory, such as performing a dance, drawing a picture, or conducting an orchestra. For these tasks, intent is likely represented at  some level in the nervous system. Nevertheless, the mechanics of the arm dictate the existence of an equilibrium whether intent is explicitly represented or not. Here our goal was to answer a more modest question, ``\textit{when} movement intent is a changing equilibrium, can we recover it despite external disturbances?''

@@

\section*{Figure Legends}
% This section is for figure legends only, do not include
% graphics in your manuscript file.
%
%\begin{figure}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure caption.  
%}
%\label{Figure_label}
%\end{figure}
\begin{figure}
\caption{
{\bf Simulated data illustrating tautology of extraction across pulse and filtered Gaussian noise disturbance types.} Intent is modeled as a minimum jerk, 5th order polynomial. Forces experienced are combined with intent via Burdet et al.'s \cite{burdet2006stability} model to produce the simulated arm trajectory. Extraction to recover intention from arm and force trajectory follows. Parameter errors are introduced into the extraction and varied according to Table 1 to estimate sensitivity. Examining the distribution of deviation in trajectories from this analysis reveals that sensitivity is not uniformly distributed throughout a reach.
}
\label{fig:synthetic}
\end{figure}

\begin{figure}[!]
\caption{
{\bf Model sensitivity testing (from the variance-based method of Saltelli et al.\cite{saltelli2010variance}) reveals that the model is mostly sensitive to stiffness.} Parameter values are varied in a Sobol-distributed fashion according to Table 1. First order sensitivity, in purple, is the error that would be removed if the parameter was fixed at its nominal value. Total sensitivity, in teal, shows error that would remain if all other parameters were fixed at their nominal values.
}
\label{fig:sensitivity}
\end{figure}

\begin{figure}[!]
\caption{
{\bf Subjects' disturbed hand trajectories and extraction of intended trajectories from them reveal that both variance and invariance of the motor plan can occur even in response to very large disturbances.} The hand path, in black, deviates from the blue baseline, as a force pulse, gray arrows, is applied. Three of the eight subjects' intent trajectories (orange) did not significantly deviate from undisturbed movement (panel A) as observed by the intent region not departing the undisturbed movement (blue) region. Shaded regions represent mean $\pm$ 95\% confidence interval on the mean. The first time at which the hand (black) region departed from undisturbed movement is marked with a green dot for each subject. Five of the eight subjects' intent trajectories did significantly deviate from undisturbed movement (panel B). The time of intent departure from the undisturbed trajectories is marked with a red dot for each subject.
}
\label{fig:anecdotes}
\end{figure}
\pagebreak
